From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-02-27 appendixB/appendixB.Rmd
WIP

Here we have the familiar demographic noise term, coupled to an environmental noise contribution which depends on the details of how the environmental noise enters the macroscopic model, $\partial_y a_1$, as well as the relative correlation times $\partial_x a_1$ and $\beta$ of the population process and environmental process.
Note that we have left this expression in terms of $\langle \eta^2 \rangle$, rather than subsituting in the equilibrium expression

>>>>

note that this process provides a justification for moving the noise from 'inside' the deterministic skeleton to merely an additive Gaussian random variable on the end of the determinstic process.  The details of how the noise term enter the dynamics are reflected in the dervative of the population equation with respect to the environmental variable, $\partial_y a_1(x,y)$, as well as the relative correlation times $\partial_x a_1$ and $\beta$ of the population process and environmental process.  This environmental contribution adds on to the variance contributed by the demographic noise.  Both demographic and environmental noise contributions are 'damped' in proportion to the stability of the macroscopic equation, $\partial_x a_1$.

Note that we have left this expression in terms of $\langle \eta^2 \rangle$, rather than subsituting in the equilibrium expression

### 2018-02-27 paper/paper.Rmd
WIP

The van Kampen expansion can also be useful in illustrating how 
environmental noise enters a model as a separate process or dimension.
Like the expansion of demographic noise, this derivation leads to 
environmental noise that is approximately Gaussian in distribution,
but unlike the canonical model, leads to explicitly autocorrelated rather 
than white noise (see derivation, Appendix B).  This result is intuitive:
for continuous time models, all fluctations are positively auto-correlated
by definition -- measure any continuous process at ever-closer intervals in
time and these measurements must converge. As the derivation makes it clear,
it is in fact the *relative* time scales of the environmental process
and population level process that really matter.  The degree of coupling
to environmental noise also depends on the derivative of the population 
dynamics with respect to the stochastically varying environmental variable:
the larger that derivative, the stronger the contribution of environmental
fluctuations to population fluctuations. For example, if extinction
rate $e$ in our Levins' model is allowed to vary
stochastically as Gaussian white noise with variance $\sigma_e^2$
and mean $\bar e$, we find the expected variance introduced in the
population to be:

@@

where $(1-\frac{\bar{e}}{c}) N$ is the equilibrium population size.
This reflects the assumption of the canonical equation, Eq \eqref{canonical}
that the standard deviation scales in direct proportion to the system size $N$,
while the demographic noise scales as the square root.  Consequently,
for a large system, the demographic contribution becomes proportionally smaller
while the environmental contribution remains the same.

This
environmental variation no doubt also arises from some more detailed
lower level process, but we need not be explicit about those dynamics
-- the expansion tells us that it is sufficient to know the overall
variance and the degree of auto-correlation.  Recent work has also
highlighted the importance of including both demographic and
environmental noise [@Melbourne2008; @Bonsall2004].  The system
size expansion still requires the assumption that the noise is small
relative to the system size.  In the case of large fluctuations,
this approximation must be replaced with a different approach, such
as the WKB approximation, as discussed in @Ovaskainen2010.

The importance of auto-correlation, has and continues to be a subject of
significant interest in the study of stochastic population dynamics
[e.g.@Roughgarden; @Ellner; @Lawton; @Ripa1999; @Marshall2015].
Early work often assumed environmental noise would be
uncorrelated as a mathematical convenience.

@@

<!--  -->

@@

>>>>

The van Kampen expansion can also be useful in illustrating how extrinsic
factors such as environmental noise enter into a stochastic model.
Like the expansion of demographic noise, this derivation leads to 
environmental noise that is approximately Gaussian in distribution,
but unlike the canonical model, leads to explicitly autocorrelated noise 
(see derivation, Appendix B).  Auto-correlated or "colored" environmental
noise has been a topic of significant theoretical interest 
[e.g. @Roughgarden; @Ellner; @Lawton; @Ripa1999; @Marshall2015], but such 
formulations start from the empirical observation of autocorrelation 
in environmental time series rather than a process-based derivation.
As the expansion clearly shows, any bounded continuous-time process will
involve auto-correlation -- measure a continuous process at ever-closer intervals in
time and these measurements must converge. As the derivation in Appendix B makes clear,
it is rather the the *relative* time scales of the environmental process
and population level process that really matter (more precisely, the relative 
magnitudes of the autocorrelation time of the environmental dynamics and population dynamics.)
An environment with comparatively short autocorrelation scale will act
effectively as white noise upon the population dynamics.

The van Kampen expansion for extrinsic factors (appendix B) also highlights
the importance of how environmental noise couples to the population dynamics.
A mechanistic description of environmental noise acts directly on parameters
of the ecological model: for instance, in our Levins' model we may imagine
that the extinction rate of patches does not occur at some fixed rate $e$
but rather at some stochastically fluctuating rate of a given mean, variance,
and autocorellation.  The system size expansion illustrates how we replace
a model where stochasticity is nested inside some parameter into the simpler
format of the canonical equation where this contribution can be merely added
on as Gaussian noise at the end of a deterministic skeleton.  (The expansion
derives that the environmental noise is proportional to the square of the derivative
of the deterministic skeleton with respect to the environmental variable,
see appendix B.) Thus for the Levin's model, we can capture the fluctuations
introduced by demographic and environmental noise at equilibrium as
a Gaussian distribution with variance given by the sum of of demographic
and environmental contributions:

@@

where $(1-\frac{\bar{e}}{c}) N$ is the equilibrium population size.
This reflects the assumption of the canonical equation, Eq \eqref{canonical}
that the standard deviation (e.g. square root of the variance) scales in direct proportion to the system size $N$,
while the demographic noise scales as the square root.  Consequently,
for a large system, the demographic contribution becomes proportionally smaller
while the environmental contribution remains the same.

These examples of how a stochastic model arises from both intrinsic and extrinsic
sources of noise through the use of the system size expansion serve both
as critique and justification of the familiar notion of a stochastic model as 
a deterministic skeleton plus some Gaussian white noise term.  Critique, because
the expansions show that both "noise" and "deterministic skeleton" are the 
result of lower-level mechanistic processes, whose precise form will 
depend on the details of those processes.  Yet the expansion also shows us that
this Gaussian noise description can be a surprisingly good approximation,
that noise intensity will converge to a stationary value whenever the system
dynamics have a stable point, and that the contribution of extrinsic
(i.e. environmental) noise factors will dominate in large systems.  These
justifications will allow us to pursue our analysis of noise-driven phenomena
using simple models in which a nonlinear deterministic skeleton is driven by
Gaussian environmental noise.

@@

<!-- Justify and contrast models to individual noise -->

@@

<!-- More frank introduction to this section, see reply to Reviewer 1-->
